---
title: "Connectome-Based Predictive Modelling of Cognitive Performance"
subtitle: "Transdiagnostic Analysis: Schizophrenia, Bipolar, ADHD & Controls"
author: "Clarisse Leong"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 12,
  fig.height = 8,
  dpi     = 300
)
```

# 1. Load Packages

```{r packages}
packages <- c("tidyverse", "ggplot2", "corrplot", "RColorBrewer",
              "viridis", "igraph", "reshape2", "patchwork",
              "ggraph", "tidygraph", "scales", "psych")

for (pkg in packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}
```

# 2. Load Data

```{r load-data}
fc_mat <- readRDS("/Users/clarisseleong/Desktop/UCLA_restFC.rds")
sc_mat <- readRDS("/Users/clarisseleong/Desktop/UCLA_SC.rds")
beh    <- read.csv("/Users/clarisseleong/Desktop/UCLA_behdata_cleaned.csv")

cat("FC dimensions:", dim(fc_mat), "\n")
cat("Beh rows:", nrow(beh), "\n")
cat("Diagnosis breakdown:\n")
print(table(beh$diagnosis))
```

# 3. Reconstruct FC Matrices

```{r reconstruct-fc}
n_regions <- 174

# Function to reconstruct symmetric matrix from upper triangle vector
vec_to_mat <- function(vec, n) {
  mat <- matrix(0, n, n)
  mat[upper.tri(mat)] <- vec
  mat <- mat + t(mat)
  diag(mat) <- 1
  return(mat)
}

# Store as 3D array: participants x regions x regions
n_sub_fc <- nrow(fc_mat)
FC_3d    <- array(0, dim = c(n_sub_fc, n_regions, n_regions))

for (i in 1:n_sub_fc) {
  FC_3d[i,,] <- vec_to_mat(fc_mat[i,], n_regions)
}

cat("FC array dimensions:", dim(FC_3d), "\n")
```

# 4. Match Participants to Behaviour

```{r match-participants}
beh_fc    <- beh[!is.na(beh$restFC), ]
n_matched <- min(nrow(beh_fc), n_sub_fc)
beh_fc    <- beh_fc[1:n_matched, ]
FC_use    <- FC_3d[1:n_matched,,]

cat("Matched N =", n_matched, "\n")
cat("Diagnosis breakdown (matched):\n")
print(table(beh_fc$diagnosis))
```

# 5. Define Cognitive Targets

```{r cognitive-targets}
cog_vars <- list(
  working_memory   = "wais_lns_totalraw",
  attention        = "cpt_fa",
  verbal_memory    = "cvlt_totcor",
  processing_speed = "colortrails_crt_time1",
  inhibition       = "stopsigl_sst_ses_ssrt"
)

# Check availability
for (nm in names(cog_vars)) {
  var <- cog_vars[[nm]]
  if (var %in% colnames(beh_fc)) {
    n_avail <- sum(!is.na(beh_fc[[var]]))
    cat(nm, "(", var, "):", n_avail, "available\n")
  } else {
    cat(nm, "(", var, "): NOT FOUND\n")
  }
}
```

# 6. Connectome-Based Predictive Modelling (CPM)

```{r}
# Residualise cognitive scores against age, sex
residualise_cog <- function(beh_df, var_name) {
  formula <- as.formula(paste(var_name, "~ age + gender"))
  fit <- lm(formula, data = beh_df, na.action = na.exclude)
  residuals(fit)
}

for (nm in names(cog_vars)) {
  var <- cog_vars[[nm]]
  if (var %in% colnames(beh_fc)) {
    beh_fc[[paste0(var, "_resid")]] <- residualise_cog(beh_fc, var)
    cog_vars[[nm]] <- paste0(var, "_resid")  # point CPM to residuals
  }
}
```

```{r cpm-function}
run_cpm <- function(FC_array, behavior, thresh = 0.01) {
  n     <- length(behavior)
  n_reg <- dim(FC_array)[2]
  pred  <- numeric(n)

  for (i in 1:n) {
    # Split train/test
    train_fc  <- FC_array[-i,,]
    train_beh <- behavior[-i]
    test_fc   <- FC_array[i,,]

    # Remove NAs from training
    valid     <- !is.na(train_beh)
    train_fc  <- train_fc[valid,,]
    train_beh <- train_beh[valid]

    # Correlate each edge with behavior in training set
    edge_r <- matrix(0, n_reg, n_reg)
    edge_p <- matrix(1, n_reg, n_reg)

    for (r1 in 1:(n_reg - 1)) {
      for (r2 in (r1 + 1):n_reg) {
        edge_vals <- train_fc[, r1, r2]
        if (sd(edge_vals, na.rm = TRUE) > 0) {
          ct <- cor.test(edge_vals, train_beh, method = "pearson")
          edge_r[r1, r2] <- ct$estimate
          edge_p[r1, r2] <- ct$p.value
        }
      }
    }

    # Select positive and negative edges
    pos_mask <- (edge_p < thresh) & (edge_r > 0)
    neg_mask <- (edge_p < thresh) & (edge_r < 0)

    # Build summary scores for test subject
    pos_sum <- sum(test_fc[pos_mask], na.rm = TRUE)
    neg_sum <- sum(test_fc[neg_mask], na.rm = TRUE)

    # Fit linear model on training data
    pos_train <- apply(train_fc, 1, function(m) sum(m[pos_mask], na.rm = TRUE))
    neg_train <- apply(train_fc, 1, function(m) sum(m[neg_mask], na.rm = TRUE))

    if (sum(pos_mask) > 0 & sum(neg_mask) > 0) {
      train_df <- data.frame(beh = train_beh, pos = pos_train, neg = neg_train)
      fit      <- lm(beh ~ pos + neg, data = train_df)
      pred[i]  <- predict(fit, newdata = data.frame(pos = pos_sum, neg = neg_sum))
    } else if (sum(pos_mask) > 0) {
      train_df <- data.frame(beh = train_beh, pos = pos_train)
      fit      <- lm(beh ~ pos, data = train_df)
      pred[i]  <- predict(fit, newdata = data.frame(pos = pos_sum))
    } else {
      pred[i]  <- mean(train_beh, na.rm = TRUE)
    }

    if (i %% 20 == 0) cat("  LOO iteration", i, "/", n, "\n")
  }

  # Compute performance
  valid_idx <- !is.na(behavior)
  r_val     <- cor(pred[valid_idx], behavior[valid_idx], use = "complete.obs")
  p_val     <- cor.test(pred[valid_idx], behavior[valid_idx])$p.value

  return(list(predicted = pred, observed = behavior,
              r = r_val, p = p_val))
}
```

```{r run-cpm}
cat("Running CPM (LOO cross-validation) — this may take 10-15 mins...\n\n")

cpm_results <- list()

for (nm in names(cog_vars)) {
  var <- cog_vars[[nm]]
  if (!var %in% colnames(beh_fc)) next

  beh_vec <- as.numeric(beh_fc[[var]])
  n_valid <- sum(!is.na(beh_vec))

  if (n_valid < 30) {
    cat("Skipping", nm, "- too few observations (", n_valid, ")\n")
    next
  }

  cat("Running CPM for:", nm, "(N =", n_valid, ")\n")
  result <- run_cpm(FC_use, beh_vec, thresh = 0.01)
  cpm_results[[nm]] <- result
  cat("  r =", round(result$r, 3), "  p =", round(result$p, 4), "\n\n")
}
```
```
```{r permutation-note}
# Permutation testing was attempted but was computationally prohibitive
# given the LOO design and sample size (N = 247).
# Conventional p-values from Pearson correlation are reported instead.
# This is noted as a limitation of the current analysis.
cat("Note: Permutation testing not run due to computational constraints.\n")
cat("Significance assessed via Pearson correlation p-values.\n")
```

# 7. Visualisation: CPM Scatter Plots

```{r cpm-scatter, fig.width=14, fig.height=10}
diag_colors <- c("CONTROL" = "#2196F3",
                 "SCHZ"    = "#F44336",
                 "BIPOLAR" = "#FF9800",
                 "ADHD"    = "#4CAF50")

plot_list <- list()

for (nm in names(cpm_results)) {
  res <- cpm_results[[nm]]
  df  <- data.frame(
    observed  = res$observed,
    predicted = res$predicted,
    diagnosis = beh_fc$diagnosis
  ) %>% filter(!is.na(observed))

  p <- ggplot(df, aes(x = observed, y = predicted, color = diagnosis)) +
    geom_point(alpha = 0.7, size = 2.5) +
    geom_smooth(method = "lm", color = "black", se = TRUE, linewidth = 1) +
    scale_color_manual(values = diag_colors) +
    theme_minimal(base_size = 12) +
    labs(
      title    = paste0("CPM: ", nm),
      subtitle = paste0("r = ", round(res$r, 3),
                        ", p = ", format.pval(res$p, digits = 3)),
      x        = "Observed Score",
      y        = "Predicted Score",
      color    = "Diagnosis"
    ) +
    theme(
      plot.title      = element_text(face = "bold", size = 13),
      plot.subtitle   = element_text(color = "gray40", size = 10),
      legend.position = "bottom"
    )

  plot_list[[nm]] <- p
}

if (length(plot_list) >= 2) {
  wrap_plots(plot_list, ncol = 2) +
    plot_annotation(
      title    = "Connectome-Based Prediction of Cognitive Performance",
      subtitle = "Resting-state FC → Cognitive scores | Leave-one-out cross-validation",
      theme    = theme(
        plot.title    = element_text(size = 16, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray40")
      )
    )
}
```

# 8. Visualisation: Group FC Matrices

```{r fc-matrices, fig.width=14, fig.height=12}
diagnoses   <- c("CONTROL", "SCHZ", "BIPOLAR", "ADHD")
diag_labels <- c("Healthy Controls", "Schizophrenia", "Bipolar", "ADHD")

n_show   <- 50
show_idx <- round(seq(1, n_regions, length.out = n_show))

fc_plots <- list()

for (d in diagnoses) {
  idx     <- which(beh_fc$diagnosis == d)
  mean_fc <- apply(FC_use[idx, show_idx, show_idx], c(2, 3), mean, na.rm = TRUE)

  df_melt <- melt(mean_fc)
  colnames(df_melt) <- c("Region1", "Region2", "FC")

  p <- ggplot(df_melt, aes(x = Region1, y = Region2, fill = FC)) +
    geom_tile() +
    scale_fill_gradientn(
      colors = c("#053061", "#2166AC", "#4393C3", "#92C5DE",
                 "#F7F7F7", "#FDDBC7", "#F4A582", "#D6604D", "#B2182B"),
      limits = c(-0.5, 0.5), oob = scales::squish,
      name   = "FC (r)"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text       = element_blank(),
      axis.ticks      = element_blank(),
      panel.grid      = element_blank(),
      plot.title      = element_text(face = "bold", size = 12, hjust = 0.5),
      legend.position = "right"
    ) +
    labs(
      title = diag_labels[which(diagnoses == d)],
      x = "Brain Region", y = "Brain Region"
    ) +
    coord_fixed()

  fc_plots[[d]] <- p
}

wrap_plots(fc_plots, ncol = 2) +
  plot_annotation(
    title    = "Mean Resting-State Functional Connectivity by Diagnosis",
    subtitle = "Top 50 regions shown | Warmer = stronger positive FC",
    theme    = theme(
      plot.title    = element_text(size = 16, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray40")
    )
  )
```

# 9. Graph Theory Metrics

```{r graph-metrics-compute}
compute_graph_metrics <- function(fc_vec, n_reg, threshold = 0.2) {
  mat <- vec_to_mat(fc_vec, n_reg)
  mat[mat < threshold] <- 0
  diag(mat) <- 0
  g <- graph_from_adjacency_matrix(mat, mode = "undirected", weighted = TRUE)

  list(
    strength   = mean(strength(g), na.rm = TRUE),
    clustering = transitivity(g, type = "average"),
    efficiency = 1 / mean_distance(g, directed = FALSE, unconnected = TRUE),
    modularity = modularity(cluster_louvain(g))
  )
}

cat("Computing graph metrics for each participant...\n")
metrics_df <- data.frame()

for (i in 1:n_matched) {
  if (i %% 30 == 0) cat("  Participant", i, "/", n_matched, "\n")
  m <- tryCatch(
    compute_graph_metrics(fc_mat[i,], n_regions),
    error = function(e) list(strength = NA, clustering = NA,
                             efficiency = NA, modularity = NA)
  )
  metrics_df <- rbind(metrics_df, data.frame(
    participant = i,
    diagnosis   = beh_fc$diagnosis[i],
    age         = beh_fc$age[i],
    gender      = beh_fc$gender[i],
    strength    = m$strength,
    clustering  = m$clustering,
    efficiency  = m$efficiency,
    modularity  = m$modularity
  ))
}
```

```{r graph-metrics-plot, fig.width=12, fig.height=10}
metrics_long <- metrics_df %>%
  select(diagnosis, strength, clustering, efficiency, modularity) %>%
  pivot_longer(-diagnosis, names_to = "metric", values_to = "value") %>%
  filter(!is.na(value))

ggplot(metrics_long, aes(x = diagnosis, y = value, fill = diagnosis)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1, aes(color = diagnosis)) +
  facet_wrap(~metric, scales = "free_y", ncol = 2,
             labeller = labeller(metric = c(
               strength   = "Node Strength",
               clustering = "Clustering Coefficient",
               efficiency = "Global Efficiency",
               modularity = "Modularity"
             ))) +
  scale_fill_manual(values  = diag_colors) +
  scale_color_manual(values = diag_colors) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text      = element_text(face = "bold", size = 12),
    axis.text.x     = element_text(angle = 30, hjust = 1, size = 10),
    legend.position = "none",
    plot.title      = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle   = element_text(size = 11, hjust = 0.5, color = "gray40")
  ) +
  labs(
    title    = "Graph Theory Metrics by Diagnosis",
    subtitle = "Resting-state functional connectome | Threshold = 0.2",
    x        = NULL,
    y        = "Value"
  )
```

# 10. Statistical Tests

```{r stats}
for (met in c("strength", "clustering", "efficiency", "modularity")) {
  cat("\n---", met, "---\n")
  formula <- as.formula(paste(met, "~ diagnosis"))
  aov_res <- aov(formula, data = metrics_df)
  print(summary(aov_res))

  if (summary(aov_res)[[1]]$`Pr(>F)`[1] < 0.05) {
    cat("Post-hoc (Tukey):\n")
    print(TukeyHSD(aov_res))
  }
}
```

# 11. Visualisation: CPM Summary

```{r cpm-summary, fig.width=10, fig.height=6}
if (length(cpm_results) > 0) {
  cpm_summary <- data.frame(
    domain = names(cpm_results),
    r      = sapply(cpm_results, function(x) x$r),
    p      = sapply(cpm_results, function(x) x$p)
  ) %>%
    mutate(
      significant  = p < 0.05,
      domain_label = c(
        working_memory   = "Working Memory",
        attention        = "Attention",
        verbal_memory    = "Verbal Memory",
        processing_speed = "Processing Speed",
        inhibition       = "Inhibitory Control"
      )[domain]
    )

  ggplot(cpm_summary,
         aes(x = reorder(domain_label, r), y = r, fill = significant)) +
    geom_col(width = 0.6, alpha = 0.85) +
    geom_text(aes(label = ifelse(significant, "*", ""),
                  y = r + 0.01), size = 6, vjust = 0) +
    geom_hline(yintercept = 0, linewidth = 0.5) +
    scale_fill_manual(
      values = c("TRUE" = "#1565C0", "FALSE" = "#90A4AE"),
      labels = c("TRUE" = "p < .05", "FALSE" = "p ≥ .05"),
      name   = "Significance"
    ) +
    coord_flip() +
    theme_minimal(base_size = 13) +
    labs(
      title    = "CPM Prediction Accuracy Across Cognitive Domains",
      subtitle = "Pearson r between predicted and observed scores | * = p < .05",
      x        = NULL,
      y        = "Prediction Accuracy (r)"
    ) +
    theme(
      plot.title      = element_text(face = "bold", size = 15, hjust = 0.5),
      plot.subtitle   = element_text(size = 10, hjust = 0.5, color = "gray40"),
      legend.position = "bottom"
    )
}
```

# 12. Save Results

```{r save}
saveRDS(cpm_results, "CPM_results.rds")
write.csv(metrics_df, "graph_metrics_per_participant.csv", row.names = FALSE)

if (length(cpm_results) > 0) {
  cpm_summary_out <- data.frame(
    domain = names(cpm_results),
    r      = sapply(cpm_results, function(x) round(x$r, 4)),
    p      = sapply(cpm_results, function(x) round(x$p, 4))
  )
  write.csv(cpm_summary_out, "CPM_summary.csv", row.names = FALSE)
}

cat("All done! Output files saved to working directory.\n")
```
